---
title: "Hands-on Ex1: Geospatial Data Wrangling with R"
---

# Overview

In this hands-on exercise, i learn how to import and wrangle geospatial data in using appropriate R packages.

## **Getting Started**

-   **sf** for importing, managing, and processing geospatial data, and

-   **tidyverse** for performing data science tasks such as importing, wrangling and visualising data.

Tidyverse consists of a family of R packages. In this hands-on exercise, the following packages will be used:

-   **readr** for importing csv data,

-   **readxl** for importing Excel worksheet,

-   **tidyr** for manipulating data,

-   **dplyr** for transforming data, and

-   **ggplot2** for visualising data

The code chunk below install and load sf and tidyverse packages into R environment.

```{r}
pacman::p_load(sf, tidyverse)
```

## **Importing Geospatial Data**

how to import the following geospatial data into R by using [*st_read()*](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package:

-   `MP14_SUBZONE_WEB_PL`, a polygon feature layer in ESRI shapefile format,

-   `CyclingPath`, a line feature layer in ESRI shapefile format, and

-   `PreSchool`, a point feature layer in kml file format.

### **Importing Polygon feature data in shapefile form**

below uses *st_read()* function of **sf** package to import `MP14_SUBZONE_WEB_PL` shapefile into R as a polygon feature data frame. Note that when the input geospatial data is in shapefile format, two arguments will be used, namely: `dsn` to define the data path and `layer` to provide the shapefile name. Also note that no extension such as .shp, .dbf, .prj and .shx are needed.

```{r}
mpsz = st_read(dsn = "data/geospatial", 
                  layer = "MP14_SUBZONE_WEB_PL")
```

we can see the 'Geometry Type' is multipolygon

### **Importing polyline feature data in shapefile form**

below uses *st_read()* function of **sf** package to import `CyclingPath` shapefile into R as line feature data frame.

```{r}
getwd()
cyclingpath = st_read(dsn = "data/geospatial", 
                         layer = "CyclingPathGazette")
```

### **Importing GIS data in kml format**

The code chunk below will be used to import the kml into R. 

```{r}
preschool = st_read("data/geospatial/PreSchoolsLocation.kml")
```

The message above reveals that `preschool` is a point feature data frame. There are a total of 1359 features and 2 fields. Different from the previous two simple feature data frame, preschool is in **wgs84** coordinates system.

to be continued - 1.5

## **Checking the Content of A Simple Feature Data Frame**

```{r}
st_geometry(mpsz)

```
